Active Learning for SAT Solver Benchmarking

### 2022-10-10 paper/paper.tex
partially improved readability of Algo 2

We transform the runtimes into discrete runtime classes which are determined on a per instance basis.
For each instance $e \in \mathcal{I}$, we use a single-link hierarchial clustering algorithm to assign the runtimes in $\bigl\{ r(a, e) \mid a \in A \bigr\}$ to one of the $k$ clusters $C_1, \dots, C_k$ such that the fastest runtimes for instance $e$ are in cluster $C_1$ and the slowest are in cluster $C_k$.
For instances with many timeouts, we have to manually fix the clustering results such that all timeouts $\tau$ are in cluster $C_k$.
%TF: maybe a bit unclear. We do not manually fix anything. There is always one cluster with only time-outs if there are any.
The runtime transformation function $\gamma_k : {\mathcal{A} \times \mathcal{I}} \rightarrow \left\lbrace 1, \dots, k \right\rbrace$ is then specified as follows.
$$\gamma_k(a, e) = j ~\Leftrightarrow~ r(a, e) \in C_j$$

Given an instance $e \in \mathcal{I}$, a solver $a \in A$ belongs to the $\gamma_k(a, e)$-fastest solvers on instance $e$. 
In our preliminary experiments, we achieved higher accuracy with prediction of such discrete runtime labels than with prediction of raw runtimes. 
The mean squared error of raw runtime prediction was within the same magnitude as the values to be predicted. 
%MI: how much more accurate are we with discrete runtime labels
Research in portfolio solvers has also shown that directly predicting runtimes without any transformation does not work well in practice~\cite{NgokoCT19,CollauttiMMO13}.

To determine solver ranks, we use the transformed runtimes $\gamma_k(a, e)$ in the adapted scoring function $s_k : A \rightarrow [1, 2 \cdot k]$ as follows.
\begin{align}
  s_k(a) := \frac{1}{|\mathcal{I}|} \sum_{e \in \mathcal{I}} \gamma'_k(a, e)
  &&
  \gamma'_k(a, e) := \begin{cases}
  	2 \cdot \gamma_k(a, e) 	& \text{if } \gamma_k(a, e) = k\\
	\gamma_k(a, e)	& \text{otherwise}
  \end{cases}
  \label{eq:rankingeq}
\end{align}

The scoring function $s_k$ induces a ranking among solvers. 
In preliminary experiments, we compared the ranking induced by $s_3$ to the ranking induced by the PAR-2 score which is commonly used in SAT competitions~\cite{FroleyksHIJS21}. 
The Spearman correlation of $s_3$ and PAR-2 ranking is about \SI{0.988} which is very close to the optimal value of 1~\cite{de2016comparing}.

That discretized runtimes are sufficient for discriminating solvers can also be shown by comparing the fraction of solver pairs whose runtimes are significantly different. 
In preliminary experiments, we used a Wilcoxon signed rank test with $\alpha = 0.05$ to compare solver pairs by their runtime vectors.
Within the SAT~Competition~2022 Anniversary track, about \SI{89.95}{\%} of solver pairs are significantly different according to that measure.
The strongly discretized runtimes $\gamma_2$ are sufficient to significantly distinguish about \SI{87.04}{\%} of solver pairs, while $\gamma_3$ is sufficient to distinguish about \SI{87.83}{\%} of solver pairs.

% MI: Enough justification, commented out the following
%For the SAT Competition~2022 Anniversary track instances, this ranking approach is capable of correctly deciding for almost all pairs of solvers which one is faster (about \SI{97.45}{\%}; $\sigma = \SI{3.68}{\%}$).
%Moreover, it decides all solvers correctly if the difference in the cluster-label-ranking scores between two solvers is at least \SI{0.11}{}.

This shows that monotonicity and score distances between ranks align quite well between both ranking methods.
The correlation between $s_k$-induced and PAR-2 induced ranks, the discriminatory power of $\gamma_k$, as well as that it enables us to use classification, show the clear advantage of runtime discretization.

@@

As already stated in the introductory section, this work addresses the \emph{New-Solver Problem} (cf.~Definition~\ref{def:new-solver-problem}).
By running Algorithm~\ref{algALBenchmark}, we obtain a prediction model $\mathcal{M}$ that provides us with a scoring function \emph{score}.
Also, it outputs the fraction of runtime that is needed to arrive at its conclusion.
First and foremost, our goal is to provide the engineer of new SAT solvers with an accurate ranking.
We define the \emph{ranking accuracy} $O_{\operatorname{acc}} \in \left[0, 1\right]$ (higher is better) by the fraction of pairs $\left(\hat{a}, a\right)$ for all $a \in \mathcal{A}$ that are decided correctly by the given ranking (cf. Algorithm~\ref{alg:eval}).
For now, we exclude the equality of solvers since data shows that the PAR-2 scores of all our solvers are different.
So, possible ranking decisions may be, for example, solver $a$ is better than $b$ or $b$ is better than $a$.
The ranking accuracy is affected by whether this decision is correctly made.
Matricon et. al.~\cite{MatriconAFSH21} also evaluate their approach by the fraction of solver pairs that is correctly decided.
Second, we also have to optimize for runtime.
The \emph{fraction of runtime} that the algorithm needs to arrive at its conclusion is denoted by $O_{\operatorname{rt}} \in \left[0, 1\right]$ (lower is better).
This metric puts the runtime summed over the sampled instances in relation to the runtime summed over all instances in the dataset (cf. Algorithm~\ref{alg:eval}).
Overall, we want to find an approach that maximizes
%
\begin{equation}
  O_\delta := \delta O_{\operatorname{acc}} + \left(1 - \delta\right) \left(1 - O_{\operatorname{rt}}\right) \enspace \textrm{,}
  \label{eq:opt}
\end{equation} 
%
whereby $\delta \in \left[0, 1\right]$ allows for linear weighting between the two optimization goals $O_{\operatorname{acc}}$ and $O_{\operatorname{rt}}$.
Plotting the approaches that maximize $O_\delta$ for all $\delta \in \left[0, 1\right]$ on a $O_{\operatorname{rt}}$-$O_{\operatorname{acc}}$-diagram provides us with a Pareto front of the best approaches for different optimization-goal weightings.

@@

  \KwIn{A pool of instances $\mathcal{I}$, a set of solvers $\mathcal{S}$ with known runtimes $r\!: \mathcal{S} \times \mathcal{I} \rightarrow \left[0, \tau\right] \cup \left\lbrace \perp \right\rbrace$, and PAR-2 scores $par2\!: \mathcal{S} \rightarrow \mathbb{R}$ for each solver.}

@@

  \For{$\hat{a} \in \mathcal{S}$}{
    $\mathcal{A} \leftarrow \mathcal{S} \setminus \left\lbrace \hat{a} \right\rbrace$

    $\left( score,\, \tilde{\mathcal{I}} \right) \leftarrow \operatorname{runALAlgorithm}\!\left(\mathcal{I}, \mathcal{A}, \left.r\right|_{\mathcal{A} \times \mathcal{I}}, \hat{a}\right)$ \tcp*{Refer to Algorithm~\ref{algALBenchmark}}

    $O_{\operatorname{acc}} \leftarrow \left(\sum_{a \in \mathcal{A}} \left[\operatorname{sign}\!\left(score\!\left(a\right) - score\!\left(\hat{a}\right)\right) = \operatorname{sign}\!\left(par2\!\left(a\right) - par2\!\left(\hat{a}\right)\right) \right]\right) / |\mathcal{A}|$

@@

    $O_{\operatorname{rt}} \leftarrow \left(\sum_{i \in \tilde{\mathcal{I}}} r\!\left(i, \hat{a}\right)\right) / \left(\sum_{i \in \mathcal{I}} r\!\left(i, \hat{a}\right)\right)$

@@

  $\left( \bar{O}_{\operatorname{acc}},\, \bar{O}_{\operatorname{rt}} \right) \leftarrow \operatorname{mean}\!\left( res \right)$

  \KwOut{Return the average ranking accuracy $\bar{O}_{\operatorname{acc}}$ and the average fraction of runtime needed $\bar{O}_{\operatorname{rt}}$.}
\end{algorithm}

>>>>

We transform the runtimes into discrete runtime labels on a per-instance basis.
For each instance $e \in \mathcal{I}$, we use a single-link hierarchical clustering algorithm to assign the runtimes in $\bigl\{ r(a, e) \mid a \in A \bigr\}$ to one of the $k$ clusters $C_1, \dots, C_k$ such that the fastest runtimes for instance $e$ are in cluster $C_1$ and the slowest are in cluster $C_k$.
For instances with many timeouts, we have to manually fix the clustering results such that all timeouts $\tau$ are in cluster $C_k$.
%TF: maybe a bit unclear. We do not manually fix anything. There is always one cluster with only time-outs if there are any.
%JB: was wondering about that as well; how do we exactly create the timeout cluster? do we only cluster with k-1, removing timeout instances, or is the timeout cluster an ordinary cluster (which might also contain on-timeout instances)?
The runtime transformation function $\gamma_k : {\mathcal{A} \times \mathcal{I}} \rightarrow \left\lbrace 1, \dots, k \right\rbrace$ is then specified as follows:
$$\gamma_k(a, e) = j ~\Leftrightarrow~ r(a, e) \in C_j$$

Given an instance $e \in \mathcal{I}$, a solver $a \in A$ belongs to the $\gamma_k(a, e)$-fastest solvers on instance $e$. 
In preliminary experiments, we achieved higher accuracy predicting such discrete runtime labels than predicting raw runtimes. 
The mean squared error of raw runtime regression was within the same magnitude as the values to be predicted. 
%MI: how much more accurate are we with discrete runtime labels
Research on portfolio solvers has also shown that directly predicting runtimes without any transformation does not work well in practice~\cite{CollauttiMMO13,NgokoCT19}.
%JB: maybe we should phrase this more defensively or just leave out this argument (might be dataset-dependend), (1) e.g., SATzilla 2008 does predict runtimes and (2) we could also have treid to predict normalized runtimes like Collautti et al. instead of discretizing (though prediction performance would probably still have been bad in our case)

To determine solver ranks, we use the transformed runtimes $\gamma_k(a, e)$ in the adapted scoring function $s_k : A \rightarrow [1, 2 \cdot k]$ as follows:
%
\begin{align}
  s_k(a) := \frac{1}{|\mathcal{I}|} \sum_{e \in \mathcal{I}} \gamma'_k(a, e)
  &&
  \gamma'_k(a, e) := \begin{cases}
  	2 \cdot \gamma_k(a, e) 	& \text{if } \gamma_k(a, e) = k\\
	\gamma_k(a, e)	& \text{otherwise}
  \end{cases}
  \label{eq:rankingeq}
\end{align}
%
I.e., we apply a PAR-2 scoring, which is commonly used in SAT competitions~\cite{FroleyksHIJS21}, on the discrete labels.
The scoring function $s_k$ induces a ranking among solvers.

However, we need to ensure that discretized labels still discriminate solvers and align with the actual PAR-2 ranking.
We analyzed the ranking induced by $s_3$ in preliminary experiments with the SAT Competition~2022 Anniversary Track~\cite{sat2022}.
According to a Wilcoxon-signed-rank test with $\alpha = 0.05$, \SI{87.83}{\%} of solver pairs have significantly different scores after discretization, only a slight drop compared to \SI{89.95}{\%} before discretization.
%While only time-out labels (solver times out or not on a particular instance) are sufficient to significantly distinguish about \SI{87.04}{\%} of solver pairs.
%\footnote{Based on hierarchical clustering with a log-single-link criterion for $k = 3$.}
Further, our ranking approach is capable of correctly deciding for almost all (about \SI{97.45}{\%}; $\sigma = \SI{3.68}{\%}$) solver pairs which solver is faster.
%Moreover, it decides all solvers correctly if the difference in the cluster-label-ranking scores between two solvers is at least \SI{0.11}{}.
%Refer to Table~\ref{tab:discreteranking} for a comparison with the PAR-2 ranking.
In particular, the Spearman correlation of $s_3$ and PAR-2 ranking is about \SI{0.988}{}, which is very close to the optimal value of 1~\cite{de2016comparing}.
All these results show that discretized runtimes are suitable for our framework.

@@

As already stated in the introductory section, this work addresses the \emph{New-Solver Problem} (cf.~Definition~\ref{def:new-solver-problem}).
We create a prediction model $\mathcal{M}$ as described in Section~\ref{sec:main-model} that provides us with a scoring function $s_k$.
% MI: Let this be the evaluation result:
%Also, it outputs the fraction of runtime that is needed to arrive at its conclusion.
First and foremost, our goal is to provide the engineer of new SAT solvers with an accurate ranking.
We define the \emph{ranking accuracy} $O_{\operatorname{acc}} \in \left[0, 1\right]$ (higher is better) by the fraction of pairs $\left(\hat{a}, a\right)$ for all $a \in \mathcal{A}$ that are decided correctly by the given ranking (cf.~Algorithm~\ref{alg:eval}).
For now, we exclude the equality of solvers since data shows that the PAR-2 scores of all our solvers are different.
So, possible ranking decisions may be, for example, solver $a$ is better than $b$ or $b$ is better than $a$.
The ranking accuracy is affected by whether this decision is correctly made.
%Matricon et. al.~\cite{MatriconAFSH21} also evaluate their approach by the fraction of solver pairs that is correctly decided.
Second, we also have to optimize for runtime.
The \emph{fraction of runtime} that the algorithm needs to arrive at its conclusion is denoted by $O_{\operatorname{rt}} \in \left[0, 1\right]$ (lower is better).
This metric puts the runtime summed over the sampled instances in relation to the runtime summed over all instances in the dataset (cf. Algorithm~\ref{alg:eval}).
Overall, we want to find an approach that maximizes
%
\begin{equation}
  O_\delta := \delta O_{\operatorname{acc}} + \left(1 - \delta\right) \left(1 - O_{\operatorname{rt}}\right) \enspace \textrm{,}
  \label{eq:opt}
\end{equation} 
%
whereby $\delta \in \left[0, 1\right]$ allows for linear weighting between the two optimization goals $O_{\operatorname{acc}}$ and $O_{\operatorname{rt}}$.
Plotting the approaches that maximize $O_\delta$ for all $\delta \in \left[0, 1\right]$ on a $O_{\operatorname{rt}}$-$O_{\operatorname{acc}}$-diagram provides us with a Pareto front of the best approaches for different optimization-goal weightings.

@@

  %\KwIn{A pool of instances $\mathcal{I}$, a set of solvers $\mathcal{S}$ with known runtimes $r\!: \mathcal{S} \times \mathcal{I} \rightarrow \left[0, \tau\right] \cup \left\lbrace \perp \right\rbrace$, and PAR-2 scores $par2\!: \mathcal{S} \rightarrow \mathbb{R}$ for each solver.}
  \KwIn{Solvers $\mathcal{A}$, Instaces $\mathcal{I}$, Runtimes $r : \mathcal{A} \times \mathcal{I} \rightarrow [0, \tau]$}
  \KwOut{Average ranking accuracy $\bar{O}_{\operatorname{acc}}$, Average fraction of runtime $\bar{O}_{\operatorname{rt}}$}
  \BlankLine

@@

  \For{$\hat{a} \in \mathcal{A}$}{
    $\mathcal{A}' \leftarrow \mathcal{S} \setminus \left\lbrace \hat{a} \right\rbrace$

    $\left( s, \tilde{\mathcal{I}} \right) \leftarrow \operatorname{runALAlgorithm}\!\left(\mathcal{I}, \mathcal{A}', r, \hat{a}\right)$ \tcp*{Refer to Algorithm~\ref{algALBenchmark}}

    $O_{\operatorname{acc}} \leftarrow \frac{1}{|\mathcal{A}|} \sum\limits_{a \in \mathcal{A}} \left[\operatorname{sign}(s(a) - s(\hat{a})) = \operatorname{sign}(par2(a) - par2(\hat{a})) \right]$

@@

	$r' \leftarrow \sum\limits_{i \in \tilde{\mathcal{I}}} r(i, \hat{a})$\;
	$r \leftarrow \sum\limits_{i \in \mathcal{I}} r(i, \hat{a})$\;
    $O_{\operatorname{rt}} \leftarrow \frac{r'}{r}$

@@

  $\left( \bar{O}_{\operatorname{acc}},\, \bar{O}_{\operatorname{rt}} \right) \leftarrow \operatorname{mean}\!\left( res \right)$ \;
  \Return $\left( \bar{O}_{\operatorname{acc}},\, \bar{O}_{\operatorname{rt}} \right)$
\end{algorithm}
